A* with Fast and Tight Heuristic in Road Networks

### 2020-08-13 paper/ch_potentials.tex
add shiny search space figure
updates and improvements to abstract and Sections 1-3

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Efficiently computing flexible and exact routes in continental-sized road networks is a tough challenge.
While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical applications.
A* is the classical approach to accelerate Dijkstra's algorithm and is used in many fields such as Experimental Algorithms, Robotics, and Artificial Intelligence.
However, A*'s performance depends on the availability of a good heuristic.
Computing a good heuristic is a challenge of its own.
To avoid the need for a heuristic, many hierarchical speedup techniques, such as Contraction Hierarchies (CH), have been developped.
These techniques are not based on A* and therefore do not need a heuristic.
They achieve speed and exactness but sacrifize A*'s flexibility.
In this paper, we use hierarchical techniques to compute tight A* heuristics.
We call our technique CH-Potential.
Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks.
With this approach, we combine some of the hierarchical speed with the flexibility offered by A* - the best of both worlds - while retaining exactness.
\end{abstract}

\section{Introduction}
\label{sec:intro}
Route planning in street networks of continental scale is a well-researched topic with many applications~\cite{bdgmpsww-rptn-16}.
It is usually formalized as a variant of the classical shortest path problem in weighted graphs.
Efficiently computing flexible and exact routes is a though challenge.
\emph{Exact} means that the computed path always has minimum weight.
\emph{Efficiently} means that the path computation is fast.
Being able to change the edge weights between path computations is \emph{flexibility}.

@@

\centering
\includegraphics[width=\columnwidth,height=200px]{fig/searchspace}

\caption{Example A* search space with CH-Potentials}
\label{img:search-space}
\end{figure}

Dijkstra's algorithm~\cite{d-ntpcg-59} achieves flexiblity and exactness but is comparatively slow.
A classic speedup approach is the well-known A* algorithm~\cite{hnr-afbhd-68}.
While flexible, A*'s performance crucially depends on the quality of the employed heuristic.
Finding a good heuristic that can be evaluated efficiently is difficult.

In this paper, we consider a setup with a preprocessing phase and a query phase.
The preprocessing phase can be slow and computes auxilary data that can be used in the query phase, which should be fast.
The input to the preprocessing phase is the graph weighted with lower bound weights $w_\ell$.
The source/origin $s$ and target/goal $t$ of the path and the actual edge weights $w_q$ are the input to the query phase.
We require that $w_q(e)\ge w_\ell(e)$ for all edges $e$.
In the query phase, the path is computed.
%
Street networks of continental scale have many millions of nodes.
While the preprocessing phase may be slow, it must eventually finish.
On graphs of this size, quadratic preprocessing time is thus prohibitive.

In this paper, we consider an algorithm, where the query phase mostly consists of an A* search.
The preprocessing phase computes auxilary data to efficiently evaluate the heuristic.
Our setup is very similar to the well-known ALT technique~\cite{gh-cspas-05,DBLP:conf/wea/DellingW07}.
We call our algorithm CH-Potentials, as it combines ideas from Contraction Hierarchies (CH)~\cite{gssv-erlrn-12} with A* heuristics, which are called potentials in~\cite{gh-cspas-05}.

Formally, a heuristic $h$ is a function that maps nodes onto numbers.
$h$ depends on the target $t$ of the search but not one the source $s$.
Without loose of generality, we assume that $h(t)=0$.
The \emph{tight} heuristic assigns to every node $v$, the length of a shortest $vt$-path.
All heuristics, for which A* is exact, are lower bounds of the tight heuristic.
CH-Potentials compute the tight heursitc with respect to $w_\ell$.
Without access to the query weights $w_q$, which are unavailable during preprocessing, no value of the heuristic can be increased, without loosing A*'s exactness.
CH-Potentials are optimal in this sense.

@@

A lot of research focuses on the inflexible setting, i.e., the special case where $w_q = w_\ell$.
In this setting, mostly hierarchical approaches dominate in termes of running time~\cite{bdgmpsww-rptn-16}.
A well-known hierarchical technique are Contraction Hierarchies (CH)~\cite{gssv-erlrn-12,dsw-cch-15}.

There is a lot of work that extends CH to handle more complex settings.
For example, in \cite{fns-opca-14,gks-rpfof-10} multi-criteria optimization is studied.
In \cite{dsw-cch-15}, CH is modified to support Live-Traffic.
Historic traffic patterns are combined with CH in \cite{swz-sfert-19,bgsv-mtdtt-13,bdpw-dtdrp-16}.
Specialized variants for electric vehicle routing are developped in \cite{bdgwz-sfpcs-19}.
While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
Further, in every setting the flexibility available at query time is fairly limited and usually encompasses only one very specific setting.
Combining all these hierarchical extensions is an unsolved problem.
%
CH-Potentials is not the first work to combine hierarchical approaches and A* \cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
However, past work mostly uses A* to accelerate hierarchical approaches even further and sacrifices A*'s flexibility in the process.
Our primary algorithm design objective is to retain A*'s flexibility.

ALT~\cite{gh-cspas-05} and CPD-Heursitics~\cite{DBLP:conf/ijcai/BonoGHS19} are the two techniques with the highest conceptual similarity to CH-Potentials.
Just as our approach, ALT is A*-based.
However, ALT's heuristic is not tight.
%
CPD-Heursitics are a combination of A* and Compressed Path Databases (CPD).
A CPD can quickly compute the first edge of a shortest path between any two nodes.
In \cite{DBLP:conf/ijcai/BonoGHS19}, SRC~\cite{DBLP:conf/socs/StrasserHB14} is used as CPD.
In every potential evaluation, a shortest path to the target is compute, whose length is used as tight heuristic.
Unfortunately, the CPD used has quadratic preprocessing running times, which is prohibitively expensive on large street networks.
%
MtsCopa~\cite{DBLP:journals/tciaig/BaierBHH15} uses a similar idea to CPD-Heuristics but is only applied in a hunter-prey application.
%
In~\cite{DBLP:conf/ijcai/0002UJAKK18} the weighted graph is embedded into Euclidean space using FastMap such that distances in space and distances in the graph roughly correspond.
The Euclidean distance is then used as A* heuristic.
%
The handling of low degree nodes in our A* search is a variation of the techniques used in TopoCore~\cite{DBLP:conf/gis/DibbeltSW15}, which is inspired by~\cite{DBLP:journals/pvldb/FunkeNS14}.

@@

To compare against related work, we introduce the concept of Oracle-A*.
It has access to a shortest distance array with respect to $w_\ell$ as potential, i.e., it is tight.
This is the tightest potential achievable during preprocessing.
The array computation time is unaccounted for, i.e., it is magically filled instantly.
Oracle-A* is not achievable in practice.
However, it is a lower bound for every algorithm following our setup.
We experimentally show that CH-Potentials are within a factor two of Oracle-A* and thus every related work.

@@

Many applications need flexibility.
In road routing, the most common ones are traffic avoidance, user preferences, and vehicle restrictions.
Traffic avoidance comes in two forms: Live~\cite{CRP,CCH} and predicted traffic~\cite{td-alt,td-ch}.
Predicted traffic is often called time-dependent routing.
%
User preferences can vary significantly~\cite{DBLP:conf/gis/FunkeS15,DBLP:conf/gis/DellingGGKTW15,DBLP:conf/gis/FunkeLS16}
For example, for some users shorter routes are more important fast routes.
Others want to avoid tunnels or highways.
%
Truck routing differs from car routing because, depending on their load, trucks are not allowed everywhere~\cite{kswz-erptd-20}.
%
Flexibility is also often used as building block in multi-agent planning settings~\cite{DBLP:journals/ai/SharonSFS15,DBLP:journals/tciaig/BaierBHH15}.
%
Electric Vehicle Routing also requires flexibility~\cite{DBLP:journals/algorithmica/BaumDPSWZ20,DBLP:conf/aaai/EisnerFS11}.

@@

We introduce CH-Potentials in Section~\ref{sec:main-algo}.
A two-phase technique to efficiently compute tight potentials.
Combined with A*, this yields an exact, efficient, and flexibile route planning algorithms.
Figure~\ref{img:search-space} depicts an example A* search with CH-Potentials.
Beside the potentials, we describe, in Section\ref{sec:low-deg-improvment}, some improvments to A* to more efficiently process low degree nodes, common in road networks.
In Section~\ref{sec:extensions}, we demonstrate CH-Potential's flexibility, by describing various extended route planning applications and showing how they can be solved using CH-Potentials.
Finally in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach.
By comparing against Oracle-A*, we show that no alternative approach can be significantly faster.

@@

PHAST \cite{dgnw-phast-13} is a CH extension that computes distances from all nodes to one target node.
First the preprocessing phase is executed analogously to the original CH.
The query phase is divided into two steps.
The first step is analogously to the CH query:
From the target node $t$, all reachable nodes via reversed down-edges are explored.
Algorithm~\ref{algo:ch-backward} shows this first step.
The second step iterates over all CH level from top to bottom.
In each iteration, all up-edges starting within the current level are followed in reverse.
After all level are processed, the distances from all nodes towards $t$ is computed.
PHAST is slightly faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
PHAST's main advantage is that is easily be parallelized.
However, we will not consider parallelization in this paper.
We refer to \cite{dgnw-phast-13} for an in-depth experimental performance analysis.
Pseudo-code is provided in Algorithm~\ref{algo:phast}.
Using PHAST, we can compute tight A* potentials in another way.
In the query phase, we first use PHAST to compute the distances from every node to $t$ with respect to the query edge weights $w_\ell$ and store the result in an array $H$.
In the next step, we run $A^*$ and implement the potential function as lookup in the array $H$.

@@

\subparagraph{Experiments}
We first investigate the effectiveness of the heuristic depending on how tight the lower bound is.
For that, we take our main graph, use the unmodified travel time metric as the lower bound $w_\ell$, and perform queries on $w_q = \alpha \cdot w_\ell$ where $\alpha$ is a factor $\geq 1$.
When $\alpha = 1$, the heuristic is as tight as possible.
With growing $\alpha$ the heuristic degrades.
Figure~\ref{fig:scaled_weights} depicts the results.
Clearly, $\alpha$ has significant influence on the running time.
Average running times range from below a millisecond to a few hundred milliseconds depending on $\alpha$.
Up to around $\alpha = 1.1$ the running time growth is roughly quadratic before leveling off to something more resembling linear growth.
We observe a significant amount of heavy outliers sometimes more than an order of magnitude slower than the median.
Generally, running times vary quite strongly even though most queries are long range queries as we draw source and target node uniformly at random.
The reason is that different queries (even with similar shortest path lengths) are affected differently by growing $\alpha$ depending on the geographic regions the shortest path traverses.
When $\alpha$ grows, more and more nodes in the area around the shortest paths are included in the search space.
So when the shortest path traverses an area where many nodes are reachable in short distance, the search space grows stronger, than when the shortest path goes only through sparse areas.
% Conclusions?

>>>>

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Efficiently computing flexible and exact routes in continental-sized road networks is a tough challenge.
While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical applications.
A* is a classical approach to accelerate Dijkstra's algorithm and is used in many fields such as Experimental Algorithms, Robotics, and Artificial Intelligence.
However, A*'s performance depends on the availability of a good heuristic.
Computing a good heuristic is a challenge of its own.
In contrast, for road networks there also exist hierarchical speedup techniques, which do not need a heuristic.
They achieve speed and exactness but sacrifice A*'s flexibility.
In this paper, we use such as Contraction Hierarchies (CH), a hierarchical technique, to compute tight A* heuristics.
We call our technique CH-Potential.
With this approach, we combine some of the speed of hierarchical techniques with the flexibility offered by A* - the best of both worlds - while retaining exactness.
Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks.
\end{abstract}

\section{Introduction}
\label{sec:intro}
Route planning in street networks of continental scale is a well-researched topic with many applications~\cite{bdgmpsww-rptn-16}.
It is usually formalized as a variant of the classical shortest path problem in weighted graphs.
Developing flexible techniques to efficiently compute exact routes is a tough challenge.
\emph{Exact} means that the computed path always has minimum weight.
\emph{Efficiently} means that the path computation is fast.
Being able to change the edge weights between path computations or handle more complex edge weights (for example travel times which change over the course of the day obtained through traffic predictions) is \emph{flexibility}.

@@

\centering
\includegraphics[width=\columnwidth]{fig/searchspace.png}

\caption{Nodes explored by A* with CH-Potentials. Nodes were explored from blue through green to red.}
\label{img:search-space}
\end{figure}

Dijkstra's algorithm~\cite{d-ntpcg-59} achieves flexibility and exactness but is comparatively slow.
Dijkstra's algorithm iteratively explores nodes of the graph with ascending distance from the origin node until the target is reached.
A classic speedup approach is the well-known A* algorithm~\cite{hnr-afbhd-68}.
It changes the order in which nodes are explored by using the distance from the origin plus an estimate of the remaining distance to the target as the ordering criterion.
Figure~\ref{img:search-space} depicts an example A* search.
While flexible, A*'s performance crucially depends on the quality of the employed estimating heuristic.
Finding a good heuristic that can be evaluated efficiently is difficult.

In this paper, we consider a setup with a preprocessing phase and a query phase.
The preprocessing phase can be slow and computes auxiliary data that can be used in the query phase, which should be fast.
The input to the preprocessing phase is the graph weighted with lower bound weights $w_\ell$.
The source/origin $s$ and target/goal $t$ of the path and the actual edge weights $w_q$ are the input to the query phase.
We require that $w_q(e)\ge w_\ell(e)$ for all edges $e$.
In the query phase, the path is computed.
%
Street networks of continental scale have many millions of nodes.
% While the preprocessing phase may be slow, it must eventually finish.
On graphs of this size, quadratic preprocessing time is prohibitive.

In this paper, we consider an algorithm, where the query phase mostly consists of an A* search.
The preprocessing phase computes auxiliary data to efficiently evaluate the heuristic.
Our setup is very similar to the well-known ALT technique~\cite{gh-cspas-05,DBLP:conf/wea/DellingW07}.
We call our algorithm CH-Potentials, as it combines ideas from Contraction Hierarchies (CH)~\cite{gssv-erlrn-12} with A* heuristics, also called potentials in~\cite{gh-cspas-05}.

Formally, a heuristic $h$ is a function that maps nodes onto numbers.
$h$ depends on the target $t$ of the search but not one the source $s$.
Without loss of generality, we assume that $h(t)=0$.
The \emph{tight} heuristic assigns to every node $v$, the length of a shortest $vt$-path.
All heuristics, for which A* is exact, are lower bounds of the tight heuristic.
CH-Potentials compute the tight heuristic with respect to $w_\ell$.
Without access to the query weights $w_q$, which are unavailable during preprocessing, no value of the heuristic can be increased, without loosing A*'s exactness.
In this sense, CH-Potentials are as tight as possible.

@@

A lot of research focuses on the inflexible setting, i.e., the special case where $w_q = w_\ell$.
In this setting, mostly hierarchical approaches dominate in terms of running time~\cite{bdgmpsww-rptn-16}.
A well-known hierarchical technique are Contraction Hierarchies (CH)~\cite{gssv-erlrn-12}.

There is a lot of work that extends CH to handle more complex settings.
For example, in~\cite{fns-opca-14,gks-rpfof-10} multi-criteria optimization is studied.
In~\cite{dsw-cch-15}, CH is modified to support Live-Traffic.
Historic traffic patterns are combined with CH in~\cite{swz-sfert-19,bgsv-mtdtt-13,bdpw-dtdrp-16}.
Specialized variants for electric vehicle routing are developed in~\cite{bdgwz-sfpcs-19}.
While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
Further, in every setting the flexibility available at query time is fairly limited and usually encompasses only one very specific setting.
Combining all these hierarchical extensions is an unsolved problem.
%
CH-Potentials is not the first work to combine hierarchical approaches and A*~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
However, past work mostly uses A* to accelerate hierarchical approaches even further and sacrifices A*'s flexibility in the process.
Our primary algorithm design objective is to retain A*'s flexibility.

ALT~\cite{gh-cspas-05} and CPD-Heuristics~\cite{DBLP:conf/ijcai/BonoGHS19} are the two techniques with the highest conceptual similarity to CH-Potentials.
Just as our approach, ALT is A*-based.
However, ALT's heuristic is not tight.
%
CPD-Heuristics are a combination of A* and Compressed Path Databases (CPD).
A CPD can quickly compute the first edge of a shortest path between any two nodes.
In~\cite{DBLP:conf/ijcai/BonoGHS19}, SRC~\cite{DBLP:conf/socs/StrasserHB14} is used as CPD.
For every distance estimation, a shortest path to the target is computed, whose length is used as the tight heuristic.
Unfortunately, the CPD used has quadratic preprocessing running times, which is prohibitively expensive on large street networks.
%
MtsCopa~\cite{DBLP:journals/tciaig/BaierBHH15} uses a similar idea to CPD-Heuristics but is only applied in a hunter-prey application.
%
In~\cite{DBLP:conf/ijcai/0002UJAKK18} the weighted graph is embedded into Euclidean space using FastMap such that distances in space and distances in the graph roughly correspond.
The Euclidean distance is then used as A* heuristic.
%
The handling of low degree nodes in our A* search is a variation of the techniques used in TopoCore~\cite{DBLP:conf/gis/DibbeltSW15}, which is inspired by~\cite{DBLP:journals/pvldb/FunkeNS14}.

@@

To compare against related work, we introduce the concept of Oracle-A*.
It has access to a shortest distance array with respect to $w_\ell$, i.e., it is tight.
This is the tightest potential achievable during preprocessing.
The array computation time is unaccounted for, i.e., it is magically filled instantly.
Oracle-A* is not achievable in practice.
However, it is a lower bound for every algorithm following our setup.
We experimentally show that CH-Potentials are within a factor two of Oracle-A* and thus every related work.

@@

Many applications need flexibility.
The most common ones are traffic avoidance, user preferences, and vehicle restrictions.
Traffic avoidance comes in two forms: Live~\cite{dgpw-crprn-13,dsw-cch-15} and predicted traffic~\cite{ndls-bastd-12,bgsv-mtdtt-13}.
Predicted traffic is often called time-dependent routing.
%
User preferences can vary significantly~\cite{DBLP:conf/gis/FunkeS15,DBLP:conf/gis/DellingGGKTW15,DBLP:conf/gis/FunkeLS16}.
For example, for some users shorter routes are more important than fast routes.
Others want to avoid tunnels or highways.
%
Truck routing differs from car routing due to night driving bans and other restrictions~\cite{kswz-erptd-20}.
%
Flexibility is also often used as building block in multi-agent planning settings~\cite{DBLP:journals/ai/SharonSFS15,DBLP:journals/tciaig/BaierBHH15}.
%
Electric Vehicle Routing also requires flexibility~\cite{DBLP:journals/algorithmica/BaumDPSWZ20,DBLP:conf/aaai/EisnerFS11}.

@@

We introduce CH-Potentials, a two-phase technique to efficiently compute tight heuristics, in Section~\ref{sec:main-algo}.
Combined with A*, this yields an exact, efficient, and flexible route planning algorithm.
Beside the heuristic, in Section\ref{sec:low-deg-improvment}, we describe some improvements to A* to more efficiently process low degree nodes, common in road networks.
In Section~\ref{sec:extensions}, we demonstrate CH-Potential's flexibility, by describing various extended route planning applications and how they can be solved using CH-Potentials.
Finally, in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach.
By comparing against Oracle-A*, we show that no alternative approach can be significantly faster, assuming that $w_\ell$ is the tightest lower bound that can be known in advance.

@@

PHAST~\cite{dgnw-phast-13} is a CH extension that computes distances from all nodes to one target node.
First the preprocessing phase is executed analogously to the original CH.
The query phase is divided into two steps.
The first step is analogously to the CH query:
From the target node $t$, all reachable nodes via reversed down-edges are explored.
Algorithm~\ref{algo:ch-backward} shows this first step.
The second step iterates over all CH level from top to bottom.
In each iteration, all up-edges starting within the current level are followed in reverse.
After all level are processed, the distances from all nodes towards $t$ is computed.
PHAST is slightly faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
PHAST's main advantage is that is easily be parallelized.
However, we will not consider parallelization in this paper.
We refer to \cite{dgnw-phast-13} for an in-depth experimental performance analysis.
Pseudo-code is provided in Algorithm~\ref{algo:phast}.
Using PHAST, we can compute tight A* potentials in another way.
In the query phase, we first use PHAST to compute the distances from every node to $t$ with respect to the query edge weights $w_\ell$ and store the result in an array $H$.
In the next step, we run $A^*$ and implement the potential function as lookup in the array $H$.

@@

\subparagraph{Experiments}
We first investigate the effectiveness of the heuristic depending on how tight the lower bound is.
For that, we take our main graph, use the unmodified travel time metric as the lower bound $w_\ell$, and perform queries on $w_q = \alpha \cdot w_\ell$ where $\alpha$ is a factor $\geq 1$.
When $\alpha = 1$, the heuristic is as tight as possible.
With growing $\alpha$ the heuristic degrades.
Figure~\ref{fig:scaled_weights} depicts the results.
Clearly, $\alpha$ has significant influence on the running time.
Average running times range from below a millisecond to a few hundred milliseconds depending on $\alpha$.
Up to around $\alpha = 1.1$ the running time growth is roughly quadratic before leveling off to something more resembling linear growth.
We observe a significant amount of heavy outliers sometimes more than an order of magnitude slower than the median.
Generally, running times vary quite strongly even though most queries are long range queries as we draw source and target node uniformly at random.
The reason is that different queries (even with similar shortest path lengths) are affected differently by growing $\alpha$ depending on the geographic regions the shortest path traverses.
When $\alpha$ grows, more and more nodes in the area around the shortest paths are explored.
So when the shortest path traverses an area where many nodes are reachable in short distance, the search space grows stronger, than when the shortest path goes only through sparse areas.
% Conclusions?
